Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's Irises

### 2021-04-29 penguins.Rmd
Minor changes before submitting (#24)
* tiny proofreading edits

* final commit before submission

 In 1935, Edgar Anderson collected size measurements for 150 flowers from three species of *Iris* on the Gaspé Peninsula in Quebec, Canada. Since then, Anderson's *Iris* observations have become a classic dataset in statistics, machine learning, and data science teaching materials. It is included in the base R **datasets** package as **iris**, making it easy for users to access without knowing much about it. However, the lack of data documentation, presence of non-intuitive variables (e.g. "sepal width"), and perfectly balanced groups with zero missing values make **iris** an inadequate and stale dataset for teaching and learning modern data science skills. Users would benefit from working with a more representative, real-world environmental dataset with a clear link to current scientific research. Importantly, Anderson’s *Iris* dataset appeared in a 1936 publication by R. A. Fisher in the *Annals of Eugenics* (which is often the first-listed citation for the dataset), inextricably linking **iris** to racist research and theory. Thus, a modern alternative to the **iris** dataset is needed. In this paper, we introduce the **palmerpenguins** R package, which includes morphological measurements collected from 2007 - 2009 for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica. The **penguins** data provides an approachable, charismatic, and near drop-in replacement for **iris** with topical relevance global climate change and environmental impacts on marine predators. Since the release on CRAN in July 2020, the **palmerpenguins** package has been downloaded over 102,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative. 
 We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling, visualization, linear modeling, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).

@@

Unlike **iris**, which contains only complete cases, the **penguins** dataset contains a small number of missing values (n~missing~ = 19, out of `r ncol(palmerpenguins::penguins) * nrow(palmerpenguins::penguins)` total values). Missing values and unequal sample sizes are common in real-world data, and adds learning opportunities to the **penguins** dataset.

@@

Both **penguins** and **iris** offer numerous opportunities to explore linear relationships and correlations, within and across species (Figures \ref{fig:penguin-pairs} & \ref{fig:iris-pairs}). A bivariate scatterplot made with the **iris** dataset reveals a clear linear relationship between petal length and petal width. Using **penguins** (\ref{fig:linear-example}), we can create a uniquely similar scatterplot with flipper length and body mass. The overall trend across all three species is approximately linear for both **iris** and **penguins**. Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted, compared to multiple linear regression with species included (\ref{fig:linear-example}).

@@

Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the \CRANpkg{palmerpenguins} R package, offer a near drop-in replacement for **iris** in a number of common use cases for data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means. In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's paradox examples. Importantly, the **penguins** dataset encompasses real-world information derived from several charismatic marine predator species with regional breeding populations notably responding to environmental change occurring throughout the Western Antarctic Peninsula region of the Southern Ocean (see @bestelmeyer_analysis_2011, @gorman_ecological_2014, @gorman_integrative_2015, @gorman_population_2017). Thus, the **penguins** dataset can facilitate discussions more broadly on biodiversity responses to global change - a contemporary and critical topic in ecology, evolution, and the environmental sciences.

>>>>

 In 1935, Edgar Anderson collected size measurements for 150 flowers from three species of *Iris* on the Gaspé Peninsula in Quebec, Canada. Since then, Anderson's *Iris* observations have become a classic dataset in statistics, machine learning, and data science teaching materials. It is included in the base R **datasets** package as **iris**, making it easy for users to access without knowing much about it. However, the lack of data documentation, presence of non-intuitive variables (e.g. "sepal width"), and perfectly balanced groups with zero missing values make **iris** an inadequate and stale dataset for teaching and learning modern data science skills. Users would benefit from working with a more representative, real-world environmental dataset with a clear link to current scientific research. Importantly, Anderson’s *Iris* dataset appeared in a 1936 publication by R. A. Fisher in the *Annals of Eugenics* (which is often the first-listed citation for the dataset), inextricably linking **iris** to racist research and theory. Thus, a modern alternative to the **iris** dataset is needed. In this paper, we introduce the **palmerpenguins** R package, which includes morphological measurements collected from 2007 - 2009 for three species of *Pygoscelis* penguins that breed on islands throughout the Palmer Archipelago, Antarctica. The **penguins** data provides an approachable, charismatic, and near drop-in replacement for **iris** with topical relevance global climate change and environmental impacts on marine predators. Since the release on CRAN in July 2020, the **palmerpenguins** package has been downloaded over 105,000 times, highlighting the demand and widespread adoption of this viable **iris** alternative. 
 We directly compare the **iris** and **penguins** datasets for selected analyses to demonstrate that R users, in particular teachers and learners currently using **iris**, can switch to the Palmer Archipelago penguins for many use cases including data wrangling, visualization, linear modeling, multivariate analysis (e.g., PCA), cluster analysis and classification (e.g., by k-means).

@@

Unlike **iris**, which contains only complete cases, the **penguins** dataset contains a small number of missing values (n~missing~ = 19, out of `r ncol(palmerpenguins::penguins) * nrow(palmerpenguins::penguins)` total values). Missing values and unequal sample sizes are common in real-world data, and create added learning opportunity to the **penguins** dataset.

@@

Both **penguins** and **iris** offer numerous opportunities to explore linear relationships and correlations, within and across species (Figures \ref{fig:penguin-pairs} & \ref{fig:iris-pairs}). A bivariate scatterplot made with the **iris** dataset reveals a clear linear relationship between petal length and petal width. Using **penguins** (\ref{fig:linear-example}), we can create a uniquely similar scatterplot with flipper length and body mass. The overall trend across all three species is approximately linear for both **iris** and **penguins**. Teachers may encourage students to explore how simple linear regression results and predictions differ when the species variable is omitted, compared to, for example, multiple linear regression with species included (\ref{fig:linear-example}).

@@

Here, we have shown that structural size measurements for Palmer Archipelago *Pygoscelis* penguins, available as **penguins** in the \CRANpkg{palmerpenguins} R package, offer a near drop-in replacement for **iris** in a number of common use cases for data science and statistics education including exploratory data visualization, linear correlation and regression, PCA, and clustering by k-means. In addition, teaching and learning opportunities in **penguins** are increased due to a greater number of variables, missing values, unequal sample sizes, and Simpson's Paradox examples. Importantly, the **penguins** dataset encompasses real-world information derived from several charismatic marine predator species with regional breeding populations notably responding to environmental change occurring throughout the Western Antarctic Peninsula region of the Southern Ocean (see @bestelmeyer_analysis_2011, @gorman_ecological_2014, @gorman_integrative_2015, @gorman_population_2017). Thus, the **penguins** dataset can facilitate discussions more broadly on biodiversity responses to global change - a contemporary and critical topic in ecology, evolution, and the environmental sciences.

### 2021-04-29 penguins.tex
fix Palmer Station LTER bibliography entries (#23)
Minor changes before submitting (#24)
* tiny proofreading edits

* final commit before submission

\abstract{%
In 1935, Edgar Anderson collected size measurements for 150 flowers from
three species of \emph{Iris} on the Gaspé Peninsula in Quebec, Canada.
Since then, Anderson's \emph{Iris} observations have become a classic
dataset in statistics, machine learning, and data science teaching
materials. It is included in the base R \textbf{datasets} package as
\textbf{iris}, making it easy for users to access without knowing much
about it. However, the lack of data documentation, presence of
non-intuitive variables (e.g.~``sepal width''), and perfectly balanced
groups with zero missing values make \textbf{iris} an inadequate and
stale dataset for teaching and learning modern data science skills.
Users would benefit from working with a more representative, real-world
environmental dataset with a clear link to current scientific research.
Importantly, Anderson's \emph{Iris} dataset appeared in a 1936
publication by R. A. Fisher in the \emph{Annals of Eugenics} (which is
often the first-listed citation for the dataset), inextricably linking
\textbf{iris} to racist research and theory. Thus, a modern alternative
to the \textbf{iris} dataset is needed. In this paper, we introduce the
\textbf{palmerpenguins} R package, which includes morphological
measurements collected from 2007 - 2009 for three species of
\emph{Pygoscelis} penguins that breed on islands throughout the Palmer
Archipelago, Antarctica. The \textbf{penguins} data provides an
approachable, charismatic, and near drop-in replacement for
\textbf{iris} with topical relevance global climate change and
environmental impacts on marine predators. Since the release on CRAN in
July 2020, the \textbf{palmerpenguins} package has been downloaded over
102,000 times, highlighting the demand and widespread adoption of this
viable \textbf{iris} alternative. We directly compare the \textbf{iris}
and \textbf{penguins} datasets for selected analyses to demonstrate that
R users, in particular teachers and learners currently using
\textbf{iris}, can switch to the Palmer Archipelago penguins for many
use cases including data wrangling, visualization, linear modeling,
multivariate analysis (e.g., PCA), cluster analysis and classification
(e.g., by k-means).
}

@@

\begin{itemize}
\tightlist
\item
  Adélie penguin data \citep{lter_structural_2020}:
  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5}{KNB-LTER
  Data Package 219.5}
\item
  Gentoo penguin data \citep{lter_structural_2020-1}:
  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5}{KNB-LTER
  Data Package 220.5}
\item
  Chinstrap penguin data \citep{lter_structural_2020-2}:
  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6}{KNB-LTER
  Data Package 221.5}
\end{itemize}

@@

Unlike \textbf{iris}, which contains only complete cases, the
\textbf{penguins} dataset contains a small number of missing values
(n\textsubscript{missing} = 19, out of 2752 total values). Missing
values and unequal sample sizes are common in real-world data, and adds
learning opportunities to the \textbf{penguins} dataset.

@@

Both \textbf{penguins} and \textbf{iris} offer numerous opportunities to
explore linear relationships and correlations, within and across species
(Figures \ref{fig:penguin-pairs} \& \ref{fig:iris-pairs}). A bivariate
scatterplot made with the \textbf{iris} dataset reveals a clear linear
relationship between petal length and petal width. Using
\textbf{penguins} (\ref{fig:linear-example}), we can create a uniquely
similar scatterplot with flipper length and body mass. The overall trend
across all three species is approximately linear for both \textbf{iris}
and \textbf{penguins}. Teachers may encourage students to explore how
simple linear regression results and predictions differ when the species
variable is omitted, compared to multiple linear regression with species
included (\ref{fig:linear-example}).

@@

Here, we have shown that structural size measurements for Palmer
Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins}
in the \CRANpkg{palmerpenguins} R package, offer a near drop-in
replacement for \textbf{iris} in a number of common use cases for data
science and statistics education including exploratory data
visualization, linear correlation and regression, PCA, and clustering by
k-means. In addition, teaching and learning opportunities in
\textbf{penguins} are increased due to a greater number of variables,
missing values, unequal sample sizes, and Simpson's paradox examples.
Importantly, the \textbf{penguins} dataset encompasses real-world
information derived from several charismatic marine predator species
with regional breeding populations notably responding to environmental
change occurring throughout the Western Antarctic Peninsula region of
the Southern Ocean (see \citet{bestelmeyer_analysis_2011},
\citet{gorman_ecological_2014}, \citet{gorman_integrative_2015},
\citet{gorman_population_2017}). Thus, the \textbf{penguins} dataset can
facilitate discussions more broadly on biodiversity responses to global
change - a contemporary and critical topic in ecology, evolution, and
the environmental sciences.

>>>>

\abstract{%
In 1935, Edgar Anderson collected size measurements for 150 flowers from
three species of \emph{Iris} on the Gaspé Peninsula in Quebec, Canada.
Since then, Anderson's \emph{Iris} observations have become a classic
dataset in statistics, machine learning, and data science teaching
materials. It is included in the base R \textbf{datasets} package as
\textbf{iris}, making it easy for users to access without knowing much
about it. However, the lack of data documentation, presence of
non-intuitive variables (e.g.~``sepal width''), and perfectly balanced
groups with zero missing values make \textbf{iris} an inadequate and
stale dataset for teaching and learning modern data science skills.
Users would benefit from working with a more representative, real-world
environmental dataset with a clear link to current scientific research.
Importantly, Anderson's \emph{Iris} dataset appeared in a 1936
publication by R. A. Fisher in the \emph{Annals of Eugenics} (which is
often the first-listed citation for the dataset), inextricably linking
\textbf{iris} to racist research and theory. Thus, a modern alternative
to the \textbf{iris} dataset is needed. In this paper, we introduce the
\textbf{palmerpenguins} R package, which includes morphological
measurements collected from 2007 - 2009 for three species of
\emph{Pygoscelis} penguins that breed on islands throughout the Palmer
Archipelago, Antarctica. The \textbf{penguins} data provides an
approachable, charismatic, and near drop-in replacement for
\textbf{iris} with topical relevance global climate change and
environmental impacts on marine predators. Since the release on CRAN in
July 2020, the \textbf{palmerpenguins} package has been downloaded over
105,000 times, highlighting the demand and widespread adoption of this
viable \textbf{iris} alternative. We directly compare the \textbf{iris}
and \textbf{penguins} datasets for selected analyses to demonstrate that
R users, in particular teachers and learners currently using
\textbf{iris}, can switch to the Palmer Archipelago penguins for many
use cases including data wrangling, visualization, linear modeling,
multivariate analysis (e.g., PCA), cluster analysis and classification
(e.g., by k-means).
}

@@

\begin{itemize}
\tightlist
\item
  Adélie penguin data \citep{lter_structural_2020}:
  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.219.5}{KNB-LTER
  Data Package 219.5}
\item
  Gentoo penguin data \citep{lter_structural_2020-1}:
  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.220.5}{KNB-LTER
  Data Package 220.5}
\item
  Chinstrap penguin data \citep{lter_structural_2020-2}:
  \href{https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-pal.221.6}{KNB-LTER
  Data Package 221.6}
\end{itemize}

@@

Unlike \textbf{iris}, which contains only complete cases, the
\textbf{penguins} dataset contains a small number of missing values
(n\textsubscript{missing} = 19, out of 2752 total values). Missing
values and unequal sample sizes are common in real-world data, and
create added learning opportunity to the \textbf{penguins} dataset.

@@

Both \textbf{penguins} and \textbf{iris} offer numerous opportunities to
explore linear relationships and correlations, within and across species
(Figures \ref{fig:penguin-pairs} \& \ref{fig:iris-pairs}). A bivariate
scatterplot made with the \textbf{iris} dataset reveals a clear linear
relationship between petal length and petal width. Using
\textbf{penguins} (\ref{fig:linear-example}), we can create a uniquely
similar scatterplot with flipper length and body mass. The overall trend
across all three species is approximately linear for both \textbf{iris}
and \textbf{penguins}. Teachers may encourage students to explore how
simple linear regression results and predictions differ when the species
variable is omitted, compared to, for example, multiple linear
regression with species included (\ref{fig:linear-example}).

@@

Here, we have shown that structural size measurements for Palmer
Archipelago \emph{Pygoscelis} penguins, available as \textbf{penguins}
in the \CRANpkg{palmerpenguins} R package, offer a near drop-in
replacement for \textbf{iris} in a number of common use cases for data
science and statistics education including exploratory data
visualization, linear correlation and regression, PCA, and clustering by
k-means. In addition, teaching and learning opportunities in
\textbf{penguins} are increased due to a greater number of variables,
missing values, unequal sample sizes, and Simpson's Paradox examples.
Importantly, the \textbf{penguins} dataset encompasses real-world
information derived from several charismatic marine predator species
with regional breeding populations notably responding to environmental
change occurring throughout the Western Antarctic Peninsula region of
the Southern Ocean (see \citet{bestelmeyer_analysis_2011},
\citet{gorman_ecological_2014}, \citet{gorman_integrative_2015},
\citet{gorman_population_2017}). Thus, the \textbf{penguins} dataset can
facilitate discussions more broadly on biodiversity responses to global
change - a contemporary and critical topic in ecology, evolution, and
the environmental sciences.
